---
title: "EBM 2024: Lecture 2"
subtitle: "Clinical Probability and Diagnostic Testing"
author: "Austin Meyer, MD, PhD, MS, MPH, MS"
date: "2025-03-14"
format:
  revealjs:
    theme: default
    html-math-method: mathjax
    css: custom.css
    slideNumber: true
    transition: slide
    title-slide-attributes:
      data-state: title-slide
    self-contained: false
---

## Roadmap - Exactly 50 Minutes...

::: {.columns}
::: {.column width="50%"}
### Fundamentals
- Probability basics – 3 minutes  
- Sensitivity & specificity – 5 minutes  
- Pretest probability – 3 minutes  
- Likelihood ratios – 4 minutes  
- Bayes' theorem – 5 minutes  
- ROC curves & cutoffs – 5 minutes
:::
::: {.column width="50%"}
### Clinical Examples
- Abdominal pain – 10 minutes  
- Respiratory infection – 7 minutes  
- Sore throat – 8 minutes  
- Q&A – 3 minutes
:::
:::

# Introduction

## From Data to Decisions

::: {.columns}
::: {.column width="50%"}
### The Clinical Question
- "Is this a pulmonary embolism?"
- "Does this child have appendicitis?"
- "Should I prescribe antibiotics?"
:::
::: {.column width="50%"}
### The Probabilistic Answer
- Moving from uncertainty to probability
- Updating beliefs with new information
- **Realize... probability is a tool for decision-making**
- **It cannot make the decision for you**
:::
:::

# Core Concepts in Clinical Probability

## Basic Probability Concepts (1/2)

- **Probability:** A measure of the likelihood of an event, ranging from 0 (impossible) to 1 (certain).
- **Conditional Probability:** The probability of an event given that another event has occurred.
- *Example:* Probability of a disease given a positive test result.

## Basic Probability Concepts (2/2)

- **Independence:** Two events are independent if the occurrence of one does not affect the probability of the other.
- **Note:** In clinical settings, many events (e.g., test results and disease presence) are not independent, requiring conditional probabilities.

## Diagnostic Test Metrics: 2x2 Table

|                           | Disease Present                   | Disease Absent                | Marginal Probability |
|---------------------------|:---------------------------------:|:-----------------------------:|:--------------------:|
| **Test Positive**         | TP                                | FP                            | *PPV = TP/(TP+FP)*   |
| **Test Negative**         | FN                                | TN                            | *NPV = TN/(FN+TN)*   |
| **Marginal Probability**  | *Sensitivity = TP/(TP+FN)*        | *Specificity = TN/(FP+TN)*    |                      |

## Diagnostic Test Metrics: Key Metrics (1/2)

- **Sensitivity:** True positive rate  
  - Formula: TP / (TP + FN)  
- **Specificity:** True negative rate  
  - Formula: TN / (TN + FP)  
- **Positive Predictive Value (PPV):** Probability of disease given a positive test  
  - Formula: TP / (TP + FP)

## Diagnostic Test Metrics: Key Metrics (2/2)

- **Negative Predictive Value (NPV):** Probability of no disease given a negative test  
  - Formula: TN / (TN + FN)  
- **Prevalence:** Proportion of the population with the disease  
  - Formula: (TP + FN) / Total  
- **Pretest Probability:** Estimated probability of disease before testing  
  - Based on prevalence and patient factors  

## Let's see how these concepts work

[Open the Predictive Value Shiny App](https://ausmeyer.shinyapps.io/predictive_value_shiny/)

## Converting Pretest to Posttest Probability: Likelihood Ratios

- **Likelihood Ratio Positive (LR+):**  
  - **Formula:** LR+ = Sensitivity / (1 - Specificity)  
  - **Interpretation:** How much more likely a positive test is in patients with the disease compared to those without.  

- **Likelihood Ratio Negative (LR-):**  
  - **Formula:** LR- = (1 - Sensitivity) / Specificity  
  - **Interpretation:** How much less likely a negative test is in patients with the disease compared to those without.  

- **Conversion Process:**  
  1. **Pretest Odds:** Pretest Odds = Pretest Probability / (1 - Pretest Probability)  
  2. **Posttest Odds:** Posttest Odds = Pretest Odds × LR  
  3. **Posttest Probability:** Posttest Probability = Posttest Odds / (1 + Posttest Odds)

## Bayes' Theorem: Posttest Probability (+Test)

- **Formula:**  
$$P(D|T+) = \frac{P(T+|D) \times P(D)}{P(T+|D) \times P(D) + P(T+|D^c) \times P(D^c)}$$ 

- **Where:**  
  - $P(D)$: Pretest probability of disease  
  - $P(T+|D)$: Sensitivity (probability of a positive test given disease)  
  - $P(T+|D^c)$: False positive rate (1 - Specificity)  
  - $P(D^c)$: 1 - Pretest probability (probability of no disease)

- **Interpretation:** The probability of having the disease given a positive test result.

## Bayes' Theorem: Posttest Probability (-Test)

- **Formula:**  
$$P(D^c|T-) = \frac{P(T-|D^c) \times P(D^c)}{P(T-|D^c) \times P(D^c) + P(T-|D) \times P(D)}$$

- **Where:**  
  - $P(D^c)$: Pretest probability of no disease (1 - P(D))  
  - $P(T-|D^c)$: Specificity (probability of a negative test given no disease)  
  - $P(T-|D)$: False negative rate (1 - Sensitivity)  
  - $P(D)$: Pretest probability of disease

- **Interpretation:** The probability of not having the disease given a negative test result.

# Comprehensive Example

## Introduction to the Dataset

**Dataset:** Pediatric Sepsis Challenge  
**Description:** Synthetic data for predicting inhospital mortality  
**Variables:**

- `inhospital_mortality`: binary outcome (1 if died, 0 if not)
- `agecalc_adm`: child's age in months
- `vaccpneumoc_adm`: self reported number of doses  
- Other variables: so many

**Note:** We will use this dataset to illustrate probability and statistical concepts in clinical decision-making for pediatrics.

## Importing and Preparing the Data

https://github.com/Kamaleswaran-Lab/The-2024-Pediatric-Sepsis-Challenge

```{r, echo=FALSE, message=FALSE}
library(dplyr)
library(ggplot2)
library(pROC)
library(knitr)

# Import the dataset
sepsis_data <- read.csv("SyntheticData_Training.csv", stringsAsFactors = FALSE) |>
  mutate(any_pneumococcal_vaccine = ifelse(vaccpneumoc_adm %in% c("3 doses", "2 doses", "1 dose"), "vaccinated", "not vaccinated"))

kable(head(sepsis_data[,1:5]))
```

## Contingency Table: Vaccination versus Mortality

```{r, echo=FALSE, message=FALSE, fig.align='center'}
# Create a contingency table for pneumococcal vaccination versus mortality
contingency_table <- table(vaccpneumococcal_vaccine = sepsis_data$any_pneumococcal_vaccine, inhospital_mortality = sepsis_data$inhospital_mortality)
kable(contingency_table)
```

- **Interpretation:** This table displays counts of children by pneumococcal vaccination status and their in hospital mortality.

## Exploratory Analysis: Age versus Inhospital Mortality

```{r, echo=FALSE, message=FALSE}
# Create a density plot
ggplot(sepsis_data, aes(x = agecalc_adm, fill = factor(inhospital_mortality))) +
  geom_density(alpha = 0.4) +
  labs(
    x = "Age (months)", 
    y = "Density",
    fill = "Inhospital Mortality",
    title = "Density Plot of Age by Inhospital Mortality"
  ) +
  scale_fill_manual(values = c("0" = "blue", "1" = "red")) +
  theme_minimal()
```

## Example of a Logistic Regression Model

```{r, fig.height=6, fig.width=8, message=FALSE, warning=FALSE}
# Set seed for reproducibility
set.seed(123)

n <- 300
x <- runif(n, min = 0, max = 10)
a <- -5
b <- 1
p <- 1 / (1 + exp(- (a + b * x)))
y <- rbinom(n, size = 1, prob = p)
synthetic_data <- data.frame(Predictor = x, Outcome = y)

ggplot(synthetic_data, aes(x = Predictor, y = Outcome)) +
  geom_jitter(height = 0.05, width = 0, alpha = 0.4) +
  stat_smooth(method = "glm", method.args = list(family = "binomial"),
              se = FALSE, color = "blue", size = 1) +
  labs(
    title = "Synthetic Data with Logistic Regression Curve",
    x = "Predictor Variable",
    y = "Binary Outcome (0/1)"
  ) +
  theme_minimal()
```

## Logistic Regression Model for Sepsis

- **Model:** `inhospital_mortality ~ sex_adm + agecalc_adm + any_pneumococcal_vaccine + hr_bpm_adm + rr_brpm_app_adm + sysbp_mmhg_adm + diasbp_mmhg_adm + temp_c_adm + muac_mm_adm + weight_kg_adm + momage_adm + glucose_mmolpl_adm`

```{r}
model_vars <- c("inhospital_mortality", "sex_adm", "agecalc_adm", "any_pneumococcal_vaccine", 
                "hr_bpm_adm", "rr_brpm_app_adm", "sysbp_mmhg_adm", "diasbp_mmhg_adm", 
                "temp_c_adm", "muac_mm_adm", "weight_kg_adm", "momage_adm", "glucose_mmolpl_adm")

sepsis_data <- sepsis_data[, model_vars] |>
  na.omit()

model <- glm(inhospital_mortality ~ sex_adm + agecalc_adm + any_pneumococcal_vaccine + hr_bpm_adm + rr_brpm_app_adm + sysbp_mmhg_adm + diasbp_mmhg_adm + temp_c_adm + muac_mm_adm + weight_kg_adm + momage_adm + glucose_mmolpl_adm, 
             data = sepsis_data, 
             family = binomial)

summary(model)
```

- **Interpretation:**
  - The model estimates the odds of inhospital mortality based on demographic, clinical, and vaccination status. Exponentiating coefficients provides the odds ratios.

## ROC Curve Analysis

```{r, fig.height=6, fig.width=7, message=FALSE, warning=FALSE}
predicted_probs <- predict(model, type = "response")
roc_obj <- roc(sepsis_data$inhospital_mortality, predicted_probs)
roc_df <- data.frame(
  specificity = rev(roc_obj$specificities),
  sensitivity = rev(roc_obj$sensitivities)
)

ggplot(roc_df, aes(x = 1 - specificity, y = sensitivity)) +
  geom_line(color = "blue", size = 1) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "red") +
  labs(
    x = "False Positive Rate (1 - Specificity)",
    y = "True Positive Rate (Sensitivity)"
  ) +
  theme_minimal()
```

- **Interpretation:** The ROC curve visualizes the model’s ability to discriminate between outcomes. The AUC summarizes overall performance.

## Optimal Cutoff using Youden Index

```{r}
best_coords <- coords(roc_obj, "best", ret = c("threshold", "sensitivity", "specificity"), transpose = FALSE)
cat("Best threshold:", best_coords[["threshold"]], "\n")
cat("Sensitivity at best threshold:", best_coords[["sensitivity"]], "\n")
cat("Specificity at best threshold:", best_coords[["specificity"]], "\n\n")

prevalence <- mean(sepsis_data$inhospital_mortality, na.rm = TRUE)
sensitivity <- best_coords[["sensitivity"]]
specificity <- best_coords[["specificity"]]
LR_plus  <- sensitivity / (1 - specificity)
LR_minus <- (1 - sensitivity) / specificity
pretest_odds <- prevalence / (1 - prevalence)
posttest_odds_pos <- pretest_odds * LR_plus
posttest_prob_pos <- posttest_odds_pos / (1 + posttest_odds_pos)
posttest_odds_neg <- pretest_odds * LR_minus
posttest_prob_neg <- posttest_odds_neg / (1 + posttest_odds_neg)

cat("Pre-test probability (positive):", round(prevalence, 2), "\n")
cat("Post-test probability (positive):", round(posttest_prob_pos, 2), "\n")
cat("Post-test probability (negative):", round(posttest_prob_neg, 2), "\n")
```

- **Interpretation:** These calculations demonstrate how the test performance shifts the probability of inhospital mortality.

## ROC Curve with Youden Index

```{r, fig.height=6, fig.width=8, message=FALSE, warning=FALSE}
youden_index <- best_coords[["sensitivity"]] + best_coords[["specificity"]] - 1
best_FPR <- 1 - best_coords[["specificity"]]
best_TPR <- best_coords[["sensitivity"]]
best_point <- data.frame(FPR = best_FPR, TPR = best_TPR)

ggplot(roc_df, aes(x = 1 - specificity, y = sensitivity)) +
  geom_line(color = "blue", size = 1) +
  geom_vline(xintercept = best_FPR, linetype = "dashed", color = "darkgreen") +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "red") +
  geom_point(data = best_point, aes(x = FPR, y = TPR), 
             color = "darkgreen", size = 3) +
  annotate("text", x = 0.2, y = 0.05, 
           label = paste0("Youden Index\nJ = Sens + Spec - 1 = ", round(youden_index, 2)),
           hjust = 0, vjust = 0, size = 5, color = "black") +
  labs(
    x = "False Positive Rate (1 - Specificity)",
    y = "True Positive Rate (Sensitivity)"
  ) +
  theme_minimal()
```

- **Interpretation:** The Youden Index identifies the optimal cutoff where sensitivity and specificity are balanced.

# Clinical Example 1

## Clinical Vignette

A 12-year-old boy presents with 24 hours of abdominal pain. The pain began around the umbilicus and **migrated to the right lower quadrant (RLQ)** over the last day. He has had **two episodes of vomiting** and reports **poor appetite**. His maximum recorded temperature is **37.5°C**. On physical examination, he exhibits **RLQ tenderness with guarding** and positive findings on **percussion and hopping tests** indicating peritoneal irritation. Laboratory studies reveal a **WBC count of 14,500/µL** with **85% neutrophils** (approximately **12,300/µL**). What should we do?

## Pediatric Appendicitis Score (PAS)

| **Criterion**                         | **Points** | **Patient Finding**          |
|---------------------------------------|:----------:|------------------------------|
| Nausea or vomiting                    | 1          | Yes                          |
| Anorexia (loss of appetite)           | 1          | Yes                          |
| Pain migration to RLQ                 | 1          | Yes                          |
| Fever (≥38°C)                         | 1          | No (37.5°C)                  |
| Cough/percussion/hopping tenderness   | 2          | Yes                          |
| RLQ tenderness                        | 2          | Yes                          |
| Leukocytosis (WBC > 10,000/µL)        | 1          | Yes (14,500/µL)              |
| Neutrophilia (ANC > 7,500/µL)         | 1          | Yes (12,300/µL)              |

**Total PAS Calculation:**

$$
\text{PAS} = 1 + 1 + 1 + 0 + 2 + 2 + 1 + 1 = 9
$$

A PAS of **7–10** is considered high risk for appendicitis.

## Ultrasound as a Diagnostic Test

- **Sensitivity:** 85%  
- **Specificity:** 93%

| **Ultrasound Result** | **If Appendicitis**  | **If No Appendicitis** |
|-----------------------|----------------------|------------------------|
| **Positive**          | 85% (True Positive)  | 7% (False Positive)    |
| **Negative**          | 15% (False Negative) | 93% (True Negative)    |

## Calculating Posttest Probability (1/2)

1. **Pretest Probability:** 50%  
2. **Pre-test Odds:**  
$$
\text{Odds} = \frac{0.5}{0.5} = 1.0
$$

3. **Likelihood Ratios:**  
- **LR+ =** $\frac{0.85}{0.07} \approx 12.1$  
- **LR– =** $\frac{0.15}{0.93} \approx 0.16$

## Calculating Posttest Probability (2/2)

- **Positive Ultrasound:**  
$$
\text{Odds}_{post} = 1.0 \times 12.1 = 12.1 \quad \Rightarrow \quad \text{Probability} \approx \frac{12.1}{13.1} \approx 92\%
$$

- **Negative Ultrasound:**  
$$
\text{Odds}_{post} = 1.0 \times 0.16 = 0.16 \quad \Rightarrow \quad \text{Probability} \approx \frac{0.16}{1.16} \approx 14\%
$$

## Clinical Decision-Making for Appendicitis

- **Positive Ultrasound (92% probability):**  
  - Immediate surgical consultation and appendectomy; initiate IV antibiotics.
- **Negative Ultrasound (14% probability):**  
  - Consider further imaging, observation, maybe abx, and maybe repeat imaging.

## Flow Diagram for Appendicitis

```{dot, fig.height=3, fig.width=7, out.width="100%", fig.margin=FALSE, fig.align='center', out.extra='style="margin: 0; padding: 0;"'}
digraph {
  graph [rankdir=LR, margin=0, pad=0.2, nodesep=0.3, ranksep=0.5, ratio=fill, size="7,2"];
  node [shape="box", style="rounded,filled", color="#3366CC", fillcolor="#E8F0FE", fontname="Helvetica", fontsize=14, margin="0.1,0.1", height=0.1, width=0.1];
  edge [fontname="Helvetica", fontsize=12];
  A [label="High Clinical Suspicion\n(PAS = 9)"];
  B [label="Perform Ultrasound"];
  C [label="Post-test Probability\n~92%"];
  D [label="Surgical Consultation &\nAppendectomy"];
  E [label="Post-test Probability\n~14%"];
  F [label="Advanced Imaging or\nObservation"];
  A -> B;
  B -> C [label="Positive"];
  C -> D;
  B -> E [label="Negative"];
  E -> F;
}
```

# Clinical Example 2

## Clinical Vignette

An 18-month-old boy, previously healthy and fully immunized, is brought to the clinic in January with a 3-day history of cough, labored breathing, and mild hypoxia (oxygen saturation 87% on room air). He attends daycare where several children have had similar symptoms. On exam, diffuse wheezing and crackles are noted, and his work of breathing is increased. There are no underlying chronic conditions. What should we do?

## Pretest Probability of Respiratory Pathogens

In winter, epidemiologic data indicate the following approximate prevalence in young children with respiratory illness:

- **Rhinovirus/Enterovirus:** 30–40%
- **Influenza:** ~10%
- **RSV:** ~5–10%
- **Adenovirus:** ~5–10%
- **Parainfluenza Viruses:** ~5–10%
- **Human Metapneumovirus:** ~5%
- **Mycoplasma pneumoniae:** ~1–2%
- **Bordetella pertussis:** <1%

*Note:* These estimates are based on published surveillance data and may vary by season and region.

## The BioFire FilmArray Respiratory Panel

The BioFire FilmArray Respiratory Panel is a multiplex PCR assay that detects a wide range of respiratory pathogens with high accuracy. Published performance characteristics from the manufacturer include (the real world accuracy is much lower):

| **Pathogen**               | **Sensitivity** | **Specificity** |
|----------------------------|-----------------|-----------------|
| Influenza A/B              | 90–100%         | 99–100%        |
| RSV                        | 95–100%         | 90–100%        |
| Rhinovirus/Enterovirus     | ~93%            | ~95%           |
| Adenovirus                 | ~89%            | ~98%           |
| Human Metapneumovirus      | ~94–95%         | ~99%           |
| Parainfluenza (1–4)        | 87–100%         | ~99%           |
| Mycoplasma pneumoniae      | ~84–100%        | ~99%           |
| Bordetella pertussis       | 95–100%         | ~99%           |

For our case, the panel returns **negative for influenza** and **positive for Mycoplasma pneumoniae**.

## Bayes’ Theorem and Post-Test Probability

We will convert our pretest probabilities to post-test probabilities using likelihood ratios.

## Influenza Calculation (1/2)

1. **Pretest Probability:** 12% (0.12)  
2. **Pretest Odds:**  
$$
\text{Odds} = \frac{0.12}{1-0.12} = \frac{0.12}{0.88} \approx 0.136
$$  
3. **Test Characteristics (Influenza):**  
   - Sensitivity = 90% (0.90)  
   - Specificity = 99% (0.99)  

## Influenza Calculation (2/2)

4. **Using Negative Test (LR–):**  
$$
LR_{-} = \frac{1-0.90}{0.99} = \frac{0.10}{0.99} \approx 0.101
$$  
5. **Post-Test Odds:**  
$$
\text{Odds}_{post} = 0.136 \times 0.101 \approx 0.0137
$$  
6. **Post-Test Probability:**  
$$
\text{Probability} = \frac{0.0137}{1+0.0137} \approx 1.35\%
$$

## Mycoplasma Calculation (1/2)

1. **Pretest Probability:** 2% (0.02)  
2. **Pretest Odds:**  
$$
\text{Odds} = \frac{0.02}{1-0.02} = \frac{0.02}{0.98} \approx 0.0204
$$  
3. **Test Characteristics (Mycoplasma):**  
   - Sensitivity = 85% (0.85)  
   - Specificity = 99% (0.99)  

## Mycoplasma Calculation (2/2)

4. **Positive Likelihood Ratio (LR+):**  
$$
LR_{+} = \frac{0.85}{1-0.99} = \frac{0.85}{0.01} = 85
$$  
5. **Post-Test Odds:**  
$$
\text{Odds}_{post} = 0.0204 \times 85 \approx 1.73
$$  
6. **Post-Test Probability:**  
$$
\text{Probability} = \frac{1.73}{1+1.73} \approx 63.4\%
$$

## Summary of Post-Test Probabilities

- **Influenza:** ~1.35% (effectively ruled out)  
- **Mycoplasma pneumoniae:** ~63.4%

## Clinical Decision-Making Based on Test Results

Given our findings:

- **Influenza:**  
  - **Post-test probability:** ~1.35%  
  - **Action:** Ruled out influenza infection.
  
- **Mycoplasma pneumoniae:**  
  - **Post-test probability:** ~63.5%  
  - **Action:** The positive result significantly raises suspicion for Mycoplasma infection.

## Take-Home Points

- **Pretest Probability Matters:**  
  - Epidemiologic data help define the likelihood of different pathogens in a given season and age group.
  
- **Test Performance is Key:**  
  - High sensitivity and specificity of the BioFire FilmArray allow for powerful shifts in probability when test results are positive.

- **Bayesian Thinking:**  
  - Use likelihood ratios to convert pretest odds to post-test odds, informing clinical decision-making.

- **Clinical Management:**  
  - In winter, a negative influenza result nearly rules out the diagnosis. A positive Mycoplasma result, although starting from a low pretest probability, warrants consideration of treatment.

# Clinical Example 3

## Clinical Vignette

A 5-year-old boy, previously healthy and fully immunized, is brought to the clinic with a 2-day history of sore throat and fever (up to 39°C). He attends daycare. On examination, he has:

- **Enlarged tonsils with exudates**
- Tender **anterior cervical lymphadenopathy**
- **Absence of cough**
 
What should we do?

## Pretest Probability of GAS Pharyngitis

In school-aged children, epidemiologic data indicate that roughly **15–30%** of sore throat cases are due to GAS. 
 
Given his age and daycare exposure, a reasonable **baseline pretest probability** is about **25%**.

## Modified Centor (McIsaac) Score

The Modified Centor Score allocates points based on clinical findings:

- **Tonsillar exudates:** +1  
- **Tender anterior cervical nodes:** +1  
- **Fever (≥38°C):** +1  
- **Absence of cough:** +1  
- **Age <15 years:** +1  

**Our Patient’s Score:**

- Exudates: **Yes** (+1)
- Tender nodes: **Yes** (+1)
- Fever: **Yes** (+1)
- No cough: **Yes** (+1)
- Age <15: **Yes** (+1)

**Total Score = 5**

A score of 5 typically increases the likelihood of GAS to roughly **50–60%**. For our purposes, we’ll assume a **post-Centor probability of 55%**.

## Rapid Antigen Detection Test (RADT)

The RADT is a quick assay for GAS with high specificity:

- **Sensitivity:** ~85%  
- **Specificity:** ~95%

From these numbers:

- **Positive Likelihood Ratio (LR+):**

  $$
  LR_{+} = \frac{0.85}{1-0.95} = \frac{0.85}{0.05} = 17
  $$

- **Negative Likelihood Ratio (LR–):**

  $$
  LR_{-} = \frac{1-0.85}{0.95} = \frac{0.15}{0.95} \approx 0.16
  $$

A positive RADT result is highly confirmatory for GAS.

# Sequential Bayesian Updates

## Step 1: Update with Centor Score (1/2)

- **Pretest Probability:** 25%  
  (Pretest Odds = 0.25/0.75 ≈ 0.33)

- **After Modified Centor Score (assume LR ~4 for a high score):**

  $$
  \text{Post-Centor Odds} = 0.33 \times 4 \approx 1.32
  $$

  $$
  \text{Post-Centor Probability} = \frac{1.32}{1 + 1.32} \approx 57\%
  $$

We round to approximately **55%** as our updated probability.

## Step 2: Update with RADT Result (2/2)

Using the **55% post-Centor probability** as the new pretest probability:

- **Pretest Odds:** 0.55/0.45 ≈ 1.22
- **If RADT is Positive (LR+ ≈ 17):**
  $$
  \text{Post-test Odds} = 1.22 \times 17 \approx 20.74
  $$
  $$
  \text{Final Post-test Probability} = \frac{20.74}{1 + 20.74} \approx 95.4\%
  $$

Thus, a positive RADT raises the probability of GAS pharyngitis to approximately **95%**.

## Sequential Testing & Clinical Decision-Making

**Sequential Testing Approach:**

1. **Modified Centor Score:**  
   - Clinical evaluation increases the probability from **25%** to about **55%**.

2. **RADT:**  
   - A positive RADT further increases the probability to **~95%**, confirming the diagnosis.

**Management:**

- **Positive RADT:**  
  - With a post-test probability of ~95%, the diagnosis is confirmed.  
  - **Action:** Start antibiotic therapy (e.g., penicillin or amoxicillin) and provide supportive care.
- **Negative RADT (if it occurred):**  
  - What would the probability be with a negative RADT?

## Take-Home Points

- **Pretest Probability is Key:**  
  - Use epidemiologic data and clinical context (e.g., daycare exposure) to estimate baseline risk.

- **Clinical Scoring Matters:**  
  - A high Modified Centor/McIsaac score substantially increases the likelihood of GAS pharyngitis.

- **Sequential Testing:**  
  - The RADT, when positive, provides strong confirmatory evidence (post-test probability ~95%).
  
- **Clinical Decision-Making:**  
  - A positive RADT in a high-risk child should prompt immediate antibiotic therapy, while a negative RADT in a high-risk case may warrant further testing with a throat culture.
  
